The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-17 paper/ms.tex
Update orbits table
Update abstract

Orbital properties of stars, computed from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to understand the
structure and evolution of the Galaxy. Stellar actions, computed from orbits,
have the attractive quality of being invariant and thus good quantitative labels
of a star's orbit. A systematic bias in the phase space position
induces 
   an orbital phase dependence
in the actions, which we interpret as a systematic error
distinct from typical observational errors and Galactic potential uncertainties.
An offset in the vertical position of 
     the Galactic midplane of
$\sim15\,\pc$ 
  for a thin disk orbit or $\sim 120\,\pc$ for a thick disk
orbit is sufficient to induce a $25\%$ systematic error in the
vertical action $J_z$. The induced error distribution is non-Gaussian and
bimodal, with neither of the modes peaking on the null value. Furthermore, we
show that the local midplane varies by $\sim100\,\pc$ at the solar circle in
Milky Way-mass cosmological zoom-in simulations from the FIRE project. From
observations that the mean vertical velocity of the Milky Way disk varies by
$\sim5\textup{--}10\,\kms$ with radius, we estimate that the true midplane
variations are of order $\sim60\textup{--}170\,\pc$, with similar offsets seen
in recent three-dimensional dust maps. Thus, current state-of-the-art action
calculations, which assume the global and local midplanes are identical, are
likely to include a systematic vertical offset depending on the volume
considered. Variation in the local standard of rest induces similar issues. The
variation of the midplane must be taken into account when performing dynamical
analysis across the large regions of the disk accessible to \textit{Gaia} and
future missions.

@@

\tablecaption{Description and names of the three orbits considered in this work.
For the last four columns: $z_{\text{max}}$ is the maximum height of the orbit,
$\frac{1}{2}(R_{\text{max}} - R_{\text{min}})$ is the magnitude of the radial
excursions of the orbit, and $\kappa$ and $\nu$ are the radial/epicyclic and
vertical frequencies of the orbit. In the epicyclic approximation, $A_z =
z_{\text{max}}$ and $A_R = \frac{1}{2}(R_{\text{max}} - R_{\text{min}})$.
\label{tab:orbits}} \tablehead{\colhead{name} & \colhead{initial position} &
\colhead{initial velocity} & \colhead{$J_R$} & \colhead{$J_{\phi}$} &
\colhead{$J_z$} & \colhead{$z_{\text{max}}$} &
\colhead{$\frac{1}{2}(R_{\text{max}} - R_{\text{min}})$} & \colhead{$\kappa$} &
\colhead{$\nu$} \\ \colhead{ } & \colhead{$\mathrm{kpc}$} &
\colhead{$\mathrm{km\,s}^{-1}$} & \colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc\,km\,s^{-1}}$} & \colhead{$\mathrm{kpc\,km\,s^{-1}}$} &
\colhead{$\mathrm{kpc}$} & \colhead{$\mathrm{kpc}$} &
\colhead{$\mathrm{Myr}^{-1}$} & \colhead{$\mathrm{Myr}^{-1}$}}
\startdata
\thin{} & (8, 0, 0) & (0, -190, 10) & 40.3 & -1520 & 0.69 & 0.12 & 1.29 & 0.049
& 0.093 \\
\thick{} & (8, 0, 0) & (0, -190, 50) & 32.5 & -1520 & 23.0 & 0.85 &
1.19 & 0.048 & 0.061 \\
\halo{} & (8, 0, 0) & (0, -190, 190) & 32.8 & -1520 &
529.1 & 6.16 & 2.34 & 0.033 & 0.025
\enddata
\end{deluxetable*}

>>>>

% Orbital properties of stars, computed from their six-dimensional phase space measurements and an assumed Galactic potential, are used to understand the structure and evolution of the Galaxy. Stellar actions, computed from orbits, are invoked as invariant and thus quantitative labels of a star’s orbit.

%   Quantitatively, an offset in the vertical position of ∼ 15 pc (∼ 120 pc) for a thin (thick) disk orbit is sufficient to induce a 25\% systematic error in the vertical action Jz. Such an offset can be trivially generated by local midplane variations. In the FIRE simulations these are on the order of ~100pc at the solar circle. For the Milky Way, from the mean vertical velocity variation of the Milky Way disk of ∼ 5–10 km/s with radius, combined with the dust-map density distribution, we estimate that the Milky Way midplane variations are of order ∼ 60–170 pc. Action calculations which assume the global and local midplanes are identical, are likely to include the induced errors we report, depending on the volume considered. Variation in the local standard of rest causes similar issues. The variation of the midplane must be taken into account when performing dynamical analysis across the large regions of the disk accessible to Gaia and future missions.

Orbital properties of stars, computed from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to understand the
structure and evolution of the Galaxy. Stellar actions, computed from orbits,
have the attractive quality of being invariant under certain assumptions and
are therefore used as quantitative labels of a star's orbit. We report a
subtle but important systematic error that is induced in the actions as a
consequence of local midplane variations expected for the Milky Way. This
error is difficult to model because it is non-Gaussian and bimodal, with
neither mode peaking on the null value. An offset in the vertical position of
the Galactic midplane of $\sim15\,\pc$ for a thin disk-like orbit or $\sim
120\,\pc$ for a thick disk-like orbit is sufficient to induce a $25\%$
systematic error in the vertical action $J_z$. In FIRE simulations of Milky
Way-mass galaxies, these variations are on the order of $\sim100\,\pc$ at the
solar circle. From observations of the mean vertical velocity variation of
$\sim5\textup{--}10\,\kms$ with radius, we estimate that the Milky Way
midplane variations are of order $\sim60\textup{--}170\,\pc$\,--\,also
consistent with recent three-dimensional dust maps. Action calculations and
orbit integrations which assume the global and local midplanes are identical
are likely to include the induced errors we report, depending on the volume
considered. Variation in the local standard of rest or distance to the
Galactic center causes similar issues. The variation of the midplane must be
taken into account when performing dynamical analysis across the large regions
of the disk accessible to \textit{Gaia} and future missions.

%  Furthermore, we
% show that the local midplane varies by $\sim100\,\pc$ at the solar circle in
% Milky Way-mass cosmological zoom-in simulations from the FIRE project. From
% observations that the mean vertical velocity of the Milky Way disk varies by
% $\sim5\textup{--}10\,\kms$ with radius, we estimate that the true midplane
% variations are of order $\sim60\textup{--}170\,\pc$, with similar offsets seen
% in recent three-dimensional dust maps. Thus, current state-of-the-art action
% calculations, which assume the global and local midplanes are identical, are
% likely to include a systematic vertical offset depending on the volume
% considered. Variation in the local standard of rest induces similar issues. The
% variation of the midplane must be taken into account when performing dynamical
% analysis across the large regions of the disk accessible to \textit{Gaia} and
% future missions.

@@

\tablecaption{Description and names of the three orbits considered in this work.
For the last four columns: $z_{\text{max}}$ is the maximum height of the orbit,
$\frac{1}{2}(R_{\text{max}} - R_{\text{min}})$ is the magnitude of the radial
excursions of the orbit, and $\kappa$ and $\nu$ are the radial/epicyclic and
vertical frequencies of the orbit. In the epicyclic approximation, $A_z =
z_{\text{max}}$ and $A_R = \frac{1}{2}(R_{\text{max}} - R_{\text{min}})$.
\label{tab:orbits}} \tablehead{\colhead{name} & \colhead{\makecell{initial \\ position}} &
\colhead{\makecell{initial \\ velocity}} & \colhead{$J_R$} & \colhead{$J_{\phi}$} &
\colhead{$J_z$} & \colhead{$z_{\text{max}}$} &
\colhead{$\frac{1}{2}(R_{\text{max}} - R_{\text{min}})$} & \colhead{$\kappa$} &
\colhead{$\nu$} \\ \colhead{ } & \colhead{$(\mathrm{kpc})$} &
\colhead{$(\mathrm{km\,s}^{-1})$} & \colhead{$(\mathrm{kpc\,km\,s^{-1}})$} &
\colhead{$(\mathrm{kpc\,km\,s^{-1}})$} & \colhead{$(\mathrm{kpc\,km\,s^{-1}})$} &
\colhead{$(\mathrm{kpc})$} & \colhead{$(\mathrm{kpc})$} &
\colhead{$(\mathrm{Myr}^{-1})$} & \colhead{$(\mathrm{Myr}^{-1})$}}
\startdata
\thin{} & (8, 0, 0) & (0, -190, 10) & 40.3 & -1520 & 0.69 & 0.12 & 1.29 & 0.049
& 0.093 \\
\thick{} & (8, 0, 0) & (0, -190, 50) & 32.5 & -1520 & 23.0 & 0.85 &
1.19 & 0.048 & 0.061 \\
\halo{} & (8, 0, 0) & (0, -190, 190) & 32.8 & -1520 &
529.1 & 6.16 & 2.34 & 0.033 & 0.025
\enddata
\end{deluxetable*}
